On the goodness-of-fit of an exponential random graph model to a social network sample

### 2021-01-20 Manuscript/manuscript.tex
Add Manuscript README

\maketitle

@@

Assume that an edge between any two nodes $X_{ij}$ can be explained either by attributes of nodes or edge patterns between nodes. Refer to the first type of predictor as an exogenous covariate and the second type as an endogenous covariate. The effect of an exogenous covariate on $X_{ij}$ can then be understood in analogy to a logistic regression model. Start by defining a set of $p$ exogenous covariates and respectively denote these as $w_{ij,1}, w_{ij,2}, ..., w_{ij,p}$. Interpret the elements of $p$ as a function of an attribute for two respective actors $i$ and $j$. An examples of such an attribute function is whether $i$ and $j$ share the same ethnic background. Another example is the difference between $i$ and $j$ on the attribute age. Since $X_{ij}$ is a dichotomous response variable, a logistic regression model can be used to obtain estimates for the unknown parameter set $\theta_{1}, \theta_{2}, ..., \theta_{p}$ such that these predict $X_{ij}$ as a function of $p$. Writing the logistic regression function in terms of the logit: \begin{equation} \text{logitPr}(X_{ij} = 1 \mid \theta) = \text{log}\frac{\text{Pr}(X_{ij} = 1 \mid \theta)}{\text{Pr}(X_{ij} = 0 \mid \theta)} = \theta_{1}w_{ij,1} + \theta_{2}w_{ij,2} + ... + \theta_{p}w_{ij, p} \end{equation} so that the probability of an edge $X_{ij}$ is given by the set of  predictors $p$ weighted by a respective set of parameters $\theta$. Positive parameter values in (1) indicate an increased probability of an edge, where negative parameter values indicate a decrease in that probability. The difference in the log-odds for two pairs of nodes $(i, j)$ and $(h, m)$ can consequently be defined as: \begin{equation} \frac{\text{logitPr}(X_{ij} = 1 \mid \theta)}{\text{logitPr}(X_{hm} = 1 \mid \theta)} = \theta_{1}(w_{ij,1} - w_{hm,1}) + \theta_{2}(w_{ij,2} - w_{hm,2}) + ... + \theta_{p}(w_{ij, p} - w_{hm, p}).\end{equation} \clearpage \noindent Assume for illustration purposes that the pairs $(i, j)$ and $(h, m)$ differ only on the node attribute of the exogenous covariate $w_{ij,1}$. In that case equation (2) reduces to $\theta_{1}$ which is equivalent to the odds-ratio. The larger the value of $\theta_{1}$, the greater the probability of the presence of an edge between the node pairs which are similar as opposed to non-similar on $w_{ij, 1}$, ceteris paribus.

@@

Define endogenous or structural covariates as the second type of predictor of the edge between any two nodes $X_{ij}$. Endogenous covariates represent counts of network configurations. A network configuration can be understood as a local sub-graph within the larger graph in which it is embedded. Examples of network configurations are counts of the triad census or the number of reciprocated vertices within a larger social network structure. The analogy with logistic regression breaks down when endogenous covariates are introduced because these variables induce dependencies among edge-variables. The issue of dependence is not extensively discussed here, but the observation is made that ignoring dependence among observations negatively affects the precision of conventional statistical models. As such, in order to make inferences about network structures a model needs to be formulated which is able to incorporate dependencies among edge-variables. Exponential random graph models (ERGMs) incorporate possible dependencies amongst edge-variables by predicting the probability of an edge conditional on what is observed in the rest of the network. More formally, the ERGM models each edge-variable $X_{ij}$ conditional on the other vertices in the network $X_{-ij}$. Write this conditional probability as $\text{Pr}(X_{ij}=1 \mid X_{−ij} = x_{−ij}, \theta)$. Concentrating on the exogenous predictors, the conditional logit becomes : \begin{equation} \text{log}\frac{\text{Pr}(X_{ij} = 1 \mid X_{-ij} = x_{-ij}, \theta)}{\text{Pr}(X_{ij} = 0 \mid X_{-ij} = x_{-ij}, \theta)} = \theta_{1}\delta_{ij,1}^{+}(x) + \theta_{2}\delta_{ij,2}^{+}(x) + ... + \theta_{p}\delta_{ij, p}^{+}(x) \end{equation} The functions $\delta_{ij,k}^{+}(x)$ in (3) are defined as change statistics for the $k$th configuration. These change statistics represent the change in transitioning from a graph for which $X_{−ij} = x_{−ij}$ and $X_{ij} = 0$ to a graph for which $X_{−ij} = x_{−ij}$ and $X_{ij} = 1$. In layman's terms,  adding any edge $X_{ij}$ to a network graph results in a possible change in any of the network configurations $p$. \clearpage \noindent This change is captured by the relevant change statistic, which when weighted by a respective parameter results in a change in the probability for $X_{ij} = 1$. As was the case for the exogenous covariates, whether an increase or decrease in the configuration count results in a respective increase or decrease in the probability for $X_{ij} = 1$ depends on the sign of the parameter estimate.  

@@

Having outlined the key fundamentals of the exponential random graph model (ERGM), write an equivalent form of the exponential random graph model (ERGM) such that it gives a probability expression for all edge-variables simultaneously. Define this as the joint form of the ERGM : \begin{equation} \text{Pr}(X = x \mid \theta) \equiv P_{\theta}(x) = \frac{1}{\kappa(\theta)}\text{exp}\{\theta_{1}z_{1}(x) + \theta_{2}z_{2}(x) + ... + \theta_{p}z_{p}(x)\}. \end{equation} The functions $z_{k}(x)$ are counts of configurations in the graph $x$, such that the corresponding change statistic for $z_{k}(x)$ is $\delta_{ij, k}^{+}(x) = z_{k}(\Delta_{ij}^{+}x) − z_{k}(\Delta_{ij}^{-}x)$, where $\Delta_{ij}^{+}x(\Delta_{ij}^{-}x)$ denotes a matrix $x$ for which $x_{ij}$ is constrained to be equal to one (zero). In analogy to what was outlined earlier, the parameters weight the relative importance of their respective configurations. The normalizing constant $\kappa(\theta) = \sum_{y \in X} \text{exp}\{\theta_{1}z_{1}(y) +$ \\ \noindent $\theta_{2}z_{2}(y)+ ... + \theta_{p}z_{p}(y)\}$ finally ensures that the sum of $P_{\theta}(x)$ over all graphs is equal to one. Put differently, its ensures hat it is a proper probability mass function. 

@@

Note that equation (4) describes a probability distribution for all graphs with $n$ nodes. Formally, the exponential random graph model (ERGM) obtains the space of all possible stochastic adjacency matrices $\textbf{X}$ for a set of nodes $n$, one of which is the empirically observed graph $x_{obs}$. This distribution of graphs over $\textbf{X}$ implies a parallel distribution of network statistics. In general, the inferential goal of the ERGM is to center this distribution of statistics over those of the observed network. Put differently, its goal is to maximize the probability of the observed graph. Define the distribution as centered on the observed values when the expected value of the networks statistics $E_{\theta}(z(X))$ is equal to the network statistics of the observed network $x_{obs}$: $E_{\theta}(z(X)) =$ \\ \noindent $z(x_{obs}))$. Equivalently, write this as the moment equation $E_{\theta}(z(X)) − z(x_{obs}) = 0$. Solving the moment equation for $\theta$ consequently provides the parameter values that provide maximal support to the observed network $x_{obs}$. \clearpage \noindent This resulting set of parameter values are also known as the maximum likelihood estimates (MLEs). 

@@

Note that the moment equation does not have an analytical solution for most exponential random graph models. If pairs of dyads in the network graph can be assumed to be independent, the maximum pseudo-likelihood estimator (MPLE) of an ERGM has been shown to equal its true likelihood \cite{hunter2008ergm}. This implies that in such cases the MLE can be found by using logistic regression as a computational vehicle \cite{hunter2008ergm}. MPLE has however been criticized because its statistical properties are not well understood when the independence assumption is violated \cite{van2009framework}. In most instances, Markov chain Monte Carlo (MCMC) methods are therefore required to obtain a solution to the moment equation. The general principle behind MCMC is to take a provisional parameter vector $\theta$ and use a Metropolis sampler to obtain a sample of graphs $x^{(1)}, x^{(2)}, ..., x^{(M)}$. In short, a Metropolis sampler generates a sequence of $m$ graphs by randomly nominating a pair of nodes $i$ and $j$ at each iteration $m$. It then proposes to update the corresponding vertex-variable such that the proposal graph $x^{*}$ at iteration $m$ is equal to $x^{(m−1)}$, except that $x_{ij}^{(m-1)}$ is set to $1 − x_{ij}^{(m−1)}$. Given the target distribution $P_{\theta}(x)$, the Metropolis sampler accepts the proposal with probability \begin{equation} \text{min} \left\{1, \frac{P_{\theta}(x^{*})}{P_{\theta}(x^{(m-1)})} \right\}. \end{equation} If the proposal is accepted, $x^{(m)} = x^{*}$; otherwise, $x^{(m)} = x^{(m−1)}$. This procedure is repeated until a sample of size $m$ which approximates the graph space $\textbf{X}$ of $P_{\theta}(x)$ is obtained. Given this sample, calculate the sample equivalent $\bar{z_{\theta}} = \frac{1}{M}(z(x^{1}) + z(x^{2}) + $ \\ \noindent $... + z(x^{M}))$ of $E_{\theta}(z(X))$, and evaluate the moment-equation $\bar{z}_{\theta} − z(x_{obs}) = 0$. If $\bar{z}_{\theta} − z(x_{obs}) \neq 0$, choose another value $\theta$ and repeat the process until a $\theta$ is found for which $\bar{z}_{\theta} − z(x_{obs}) = 0$, the MLE. 

@@

A number of MCMC algorithms exist for obtaining the MLE. The method used here was originally proposed by \citeA{geyer1992constrained}. \clearpage \noindent This approach solves the moment equation by representing the graph space $\textbf{X}$ through a fixed sample of graphs for a provisional value of the parameter vector $\theta$. As outlined previously, such a fixed sample is obtained from the model $P_{\tilde{\theta}}(x)$ with a Metropolis sampler. The method then applies Fisher scoring to find $\theta$ such that $\bar{z}_\theta − z(x_{obs}) = 0$. Since the sample of graphs is an approximation of $\textbf{X}$, a weighted average of the networks statistics is used to determine $\bar{z_{\theta}}$. The sample average $\bar{f}_{\theta} = w^{(1)}f(x^{(1)}) + w^{(2)}f(x^{(2)}) + ... + w^{(M)}f(x^{(M)})$, of a function $f(x)$, with weights \begin{equation} w^{m} = \frac{e^{(\theta_{1} - \tilde{\theta_{1}})z_{1}(x^{(m)}) + (\theta_{2} - \tilde{\theta_{2}})z_{2}(x^{(m)}) + ... + (\theta_{p} - \tilde{\theta_{p}})z_{p}(x^{(m)})}}{\sum_{k=1}^{M}e^{(\theta_{1} - \tilde{\theta_{1}})z_{1}(x^{(m)}) + (\theta_{2} - \tilde{\theta_{2}})z_{2}(x^{(m)}) + ... + (\theta_{p} - \tilde{\theta_{p}})z_{p}(x^{(m)})}} \end{equation} is an asymptotic approximation to the true expected value $E_{\theta}(f(X))$ when $\theta \approx \tilde{\theta}$. To solve the moment equation, a sequence of parameters $\theta^{(0)} = \tilde{\theta}, \theta^{(1)}, \theta^{(2)}, ..., \theta^{(g)}$ is generated using Fisher scoring. The updating step of $\theta^{(g)}$ in the Fisher scoring algorithm is of the form: \begin{equation} \theta^{(g)} = \theta^{(g-1)} - \frac{1}{D(\theta^{(g-1)})} \left\{\sum_{m=1}^{M}w^{m}z(x^{m}) - z(x_{obs}) \right\} \end{equation} and iterates until that $\theta^{(g-1)}$ for which $\sum_{m=1}^{M}w^{m}z(x^{m}) = E_{\theta}(z(X))$ such that $E_{\theta}(z(X)) - z(x_{obs}) = 0$, which is the MLE. The term $D(\theta)$ in the Fisher scoring algorithm is a scaling matrix for the difference between observed values and the expected values of the network statistics. In short, the network statistics need to be scaled because they differ in their sensitivity to changes in the parameter vector $\theta$. The matrix $D(\theta)$ is set to the weighted sample covariance $\sum_{m}w^{(m)}z(x^{(m)})z(x^{(m)})^{T} - $ \\ \noindent $\left[ \sum_{m}w^{(m)}z(x^{(m)}) \right] \left[ \sum_{m}w^{(m)}z(x^{(m)}) \right]^{T}$. 

@@

Having introduced the exponential random graph model (ERGM) for a single observed social network, let us now define the ERGM at the level of the network sample. One method for obtaining an ERGM over a network sample is a meta-analysis \cite{lubbers2003group}. The idea behind a meta-analysis is to use a weighted average to summarize over the results of multiple statistical models, here ERGMs \cite{lubbers2003group}. To illustrate the method, let us imagine a scenario where we want to apply an ERGM to a network sample of size $\{2, ..., m\}$. \clearpage \noindent Initially, an ERGM is fitted to each network in the sample, resulting in a set of ERGMs with size equal to $m$, each with its own set of parameters and standard errors. In order to summarize these separate ERGMs, a meta-analytic approach as outlined in \citeA{lubbers2003group} can be applied. This method splits the coefficients of the separate ERGMs into an average coefficient and a network-dependent deviation \cite{lubbers2003group}. Let $\theta$ denote a parameter included in each ERGM. The meta-level regression equation for this parameter can then be written as: \begin{equation} \hat{\theta} = \mu_{\theta} + U_{m} + E_{m} \end{equation} where $\hat{\theta}$ denotes the estimated parameter value for network $m$, $\mu_{\theta}$ is the average parameter coefficient in the population, $U_{m}$ is the true deviation of network $m$, which has a mean of 0 and a variance $\sigma^{2}_{\theta}$, and $E_{m}$ denotes the estimation error associated with the true parameter value $\theta_{m}$. The variance of $E_{m}$ is assumed to be known and can be obtained as the squared standard error of $\theta_{m}$ \cite{lubbers2003group}. The meta-analysis consequently seeks to differentiate between the true and error variance, in order to obtain more precise estimators for $\mu_{\theta}$ and $\sigma^{2}_{\theta}$. Put differently, by differentiating between true and error variances the method more accurately weights $\mu_{\theta}$ based on differences in standard errors between networks. \citeA{lubbers2003group} suggests using iteratively re-weighted least squares (IRLS) for estimating (8), such that networks with large standard errors have less influence on $\mu_{\theta}$ than networks with small standard errors. This results in a weighted vector of parameter means, each with an associated standard error. These results can then be interpreted as representing a general ERGM description of the network sample. The reader is referred to \citeA{lubbers2003group} for a complete overview and practical example of the meta-analytic procedure.

@@

\clearpage As stated earlier, the information criteria formulated by \citeA{akaike1974new} - Aikake's information criterion (AIC) - and \cite{schwarz1978estimating} - Schwarz's information criterion (SIC) - will here be used to quantify GOF. The AIC and SIC can be defined respectively as \begin{equation} \text{AIC(M)} = -2 \text{log} P_{\hat{\theta}} + 2p, \end{equation} \begin{equation} \text{SIC(M)} = -2\text{log}P_{\hat{\theta}} + \text{log}S. \end{equation} Here, M refers to a particular ERGM, $P_{\bar{\theta}}$ refers to the maximized likelihood under M, $p$ refers to the number of parameters in the ERGM, and $S$ refers to the size of the sample \cite{wit2012all}. For a derivation of these two formulae with respect to the Kullback–Leibler divergence the reader is referred to \citeA{wit2012all}. 

>>>>

\thispagestyle{empty}
\maketitle
\pagenumbering{arabic}

@@

Assume that an edge between any two nodes $X_{ij}$ can be explained either by attributes of nodes or edge patterns between nodes. Refer to the first type of predictor as an exogenous covariate and the second type as an endogenous covariate. The effect of an exogenous covariate on $X_{ij}$ can then be understood in analogy to a logistic regression model. Start by defining a set of $p$ exogenous covariates and respectively denote these as $w_{ij,1}, w_{ij,2}, ..., w_{ij,p}$. Interpret the elements of $p$ as a function of an attribute for two respective actors $i$ and $j$. An examples of such an attribute function is whether $i$ and $j$ share the same ethnic background. Another example is the difference between $i$ and $j$ on the attribute age. Since $X_{ij}$ is a dichotomous response variable, a logistic regression model can be used to obtain estimates for the unknown parameter set $\theta_{1}, \theta_{2}, ..., \theta_{p}$ such that these predict $X_{ij}$ as a function of $p$. Writing the logistic regression function in terms of the logit: \begin{equation} \label{eqn:eqn1} \text{logitPr}(X_{ij} = 1 \mid \theta) = \text{log}\frac{\text{Pr}(X_{ij} = 1 \mid \theta)}{\text{Pr}(X_{ij} = 0 \mid \theta)} = \theta_{1}w_{ij,1} + \theta_{2}w_{ij,2} + ... + \theta_{p}w_{ij, p} \end{equation} so that the probability of an edge $X_{ij}$ is given by the set of  predictors $p$ weighted by a respective set of parameters $\theta$. Positive parameter values in \eqref{eqn:eqn1} indicate an increased probability of an edge, where negative parameter values indicate a decrease in that probability. The difference in the log-odds for two pairs of nodes $(i, j)$ and $(h, m)$ can consequently be defined as: \begin{equation} \label{eqn:eqn2} \frac{\text{logitPr}(X_{ij} = 1 \mid \theta)}{\text{logitPr}(X_{hm} = 1 \mid \theta)} = \theta_{1}(w_{ij,1} - w_{hm,1}) + \theta_{2}(w_{ij,2} - w_{hm,2}) + ... + \theta_{p}(w_{ij, p} - w_{hm, p}).\end{equation} \clearpage \noindent Assume for illustration purposes that the pairs $(i, j)$ and $(h, m)$ differ only on the node attribute of the exogenous covariate $w_{ij,1}$. In that case equation \eqref{eqn:eqn2} reduces to $\theta_{1}$ which is equivalent to the odds-ratio. The larger the value of $\theta_{1}$, the greater the probability of the presence of an edge between the node pairs which are similar as opposed to non-similar on $w_{ij, 1}$, ceteris paribus.

@@

Define endogenous or structural covariates as the second type of predictor of the edge between any two nodes $X_{ij}$. Endogenous covariates represent counts of network configurations. A network configuration can be understood as a local sub-graph within the larger graph in which it is embedded. Examples of network configurations are counts of the triad census or the number of reciprocated vertices within a larger social network structure. The analogy with logistic regression breaks down when endogenous covariates are introduced because these variables induce dependencies among edge-variables. The issue of dependence is not extensively discussed here, but the observation is made that ignoring dependence among observations negatively affects the precision of conventional statistical models. As such, in order to make inferences about network structures a model needs to be formulated which is able to incorporate dependencies among edge-variables. Exponential random graph models (ERGMs) incorporate possible dependencies amongst edge-variables by predicting the probability of an edge conditional on what is observed in the rest of the network. More formally, the ERGM models each edge-variable $X_{ij}$ conditional on the other vertices in the network $X_{-ij}$. Write this conditional probability as $\text{Pr}(X_{ij}=1 \mid X_{−ij} = x_{−ij}, \theta)$. Concentrating on the exogenous predictors, the conditional logit becomes : \begin{equation} \label{eqn:eqn3} \text{log}\frac{\text{Pr}(X_{ij} = 1 \mid X_{-ij} = x_{-ij}, \theta)}{\text{Pr}(X_{ij} = 0 \mid X_{-ij} = x_{-ij}, \theta)} = \theta_{1}\delta_{ij,1}^{+}(x) + \theta_{2}\delta_{ij,2}^{+}(x) + ... + \theta_{p}\delta_{ij, p}^{+}(x) \end{equation} The functions $\delta_{ij,k}^{+}(x)$ in \eqref{eqn:eqn3} are defined as change statistics for the $k$th configuration. These change statistics represent the change in transitioning from a graph for which $X_{−ij} = x_{−ij}$ and $X_{ij} = 0$ to a graph for which $X_{−ij} = x_{−ij}$ and $X_{ij} = 1$. In layman's terms,  adding any edge $X_{ij}$ to a network graph results in a possible change in any of the network configurations $p$. \clearpage \noindent This change is captured by the relevant change statistic, which when weighted by a respective parameter results in a change in the probability for $X_{ij} = 1$. As was the case for the exogenous covariates, whether an increase or decrease in the configuration count results in a respective increase or decrease in the probability for $X_{ij} = 1$ depends on the sign of the parameter estimate.  

@@

Having outlined the key fundamentals of the exponential random graph model (ERGM), write an equivalent form of the exponential random graph model (ERGM) such that it gives a probability expression for all edge-variables simultaneously. Define this as the joint form of the ERGM : \begin{equation} \label{eqn:eqn4} \text{Pr}(X = x \mid \theta) \equiv P_{\theta}(x) = \frac{1}{\kappa(\theta)}\text{exp}\{\theta_{1}z_{1}(x) + \theta_{2}z_{2}(x) + ... + \theta_{p}z_{p}(x)\}. \end{equation} The functions $z_{k}(x)$ are counts of configurations in the graph $x$, such that the corresponding change statistic for $z_{k}(x)$ is $\delta_{ij, k}^{+}(x) = z_{k}(\Delta_{ij}^{+}x) − z_{k}(\Delta_{ij}^{-}x)$, where $\Delta_{ij}^{+}x(\Delta_{ij}^{-}x)$ denotes a matrix $x$ for which $x_{ij}$ is constrained to be equal to one (zero). In analogy to what was outlined earlier, the parameters weight the relative importance of their respective configurations. The normalizing constant $\kappa(\theta) = \sum_{y \in X} \text{exp}\{\theta_{1}z_{1}(y) +$ \\ \noindent $\theta_{2}z_{2}(y)+ ... + \theta_{p}z_{p}(y)\}$ finally ensures that the sum of $P_{\theta}(x)$ over all graphs is equal to one. Put differently, its ensures hat it is a proper probability mass function. 

@@

Note that equation \eqref{eqn:eqn4} describes a probability distribution for all graphs with $n$ nodes. Formally, the exponential random graph model (ERGM) obtains the space of all possible stochastic adjacency matrices $\textbf{X}$ for a set of nodes $n$, one of which is the empirically observed graph $x_{obs}$. This distribution of graphs over $\textbf{X}$ implies a parallel distribution of network statistics. In general, the inferential goal of the ERGM is to center this distribution of statistics over those of the observed network. Put differently, its goal is to maximize the probability of the observed graph. Define the distribution as centered on the observed values when the expected value of the networks statistics $E_{\theta}(z(X))$ is equal to the network statistics of the observed network $x_{obs}$: $E_{\theta}(z(X)) =$ \\ \noindent $z(x_{obs}))$. Equivalently, write this as the moment equation $E_{\theta}(z(X)) − z(x_{obs}) = 0$. Solving the moment equation for $\theta$ consequently provides the parameter values that provide maximal support to the observed network $x_{obs}$. \clearpage \noindent This resulting set of parameter values are also known as the maximum likelihood estimates (MLEs). 

@@

Note that the moment equation does not have an analytical solution for most exponential random graph models. If pairs of dyads in the network graph can be assumed to be independent, the maximum pseudo-likelihood estimator (MPLE) of an ERGM has been shown to equal its true likelihood \cite{hunter2008ergm}. This implies that in such cases the MLE can be found by using logistic regression as a computational vehicle \cite{hunter2008ergm}. MPLE has however been criticized because its statistical properties are not well understood when the independence assumption is violated \cite{van2009framework}. In most instances, Markov chain Monte Carlo (MCMC) methods are therefore required to obtain a solution to the moment equation. The general principle behind MCMC is to take a provisional parameter vector $\theta$ and use a Metropolis sampler to obtain a sample of graphs $x^{(1)}, x^{(2)}, ..., x^{(M)}$. In short, a Metropolis sampler generates a sequence of $m$ graphs by randomly nominating a pair of nodes $i$ and $j$ at each iteration $m$. It then proposes to update the corresponding vertex-variable such that the proposal graph $x^{*}$ at iteration $m$ is equal to $x^{(m−1)}$, except that $x_{ij}^{(m-1)}$ is set to $1 − x_{ij}^{(m−1)}$. Given the target distribution $P_{\theta}(x)$, the Metropolis sampler accepts the proposal with probability \begin{equation} \label{eqn:eqn5} \text{min} \left\{1, \frac{P_{\theta}(x^{*})}{P_{\theta}(x^{(m-1)})} \right\}. \end{equation} If the proposal is accepted, $x^{(m)} = x^{*}$; otherwise, $x^{(m)} = x^{(m−1)}$. This procedure is repeated until a sample of size $m$ which approximates the graph space $\textbf{X}$ of $P_{\theta}(x)$ is obtained. Given this sample, calculate the sample equivalent $\bar{z_{\theta}} = \frac{1}{M}(z(x^{1}) + z(x^{2}) + $ \\ \noindent $... + z(x^{M}))$ of $E_{\theta}(z(X))$, and evaluate the moment-equation $\bar{z}_{\theta} − z(x_{obs}) = 0$. If $\bar{z}_{\theta} − z(x_{obs}) \neq 0$, choose another value $\theta$ and repeat the process until a $\theta$ is found for which $\bar{z}_{\theta} − z(x_{obs}) = 0$, the MLE. 

@@

A number of MCMC algorithms exist for obtaining the MLE. The method used here was originally proposed by \citeA{geyer1992constrained}. \clearpage \noindent This approach solves the moment equation by representing the graph space $\textbf{X}$ through a fixed sample of graphs for a provisional value of the parameter vector $\theta$. As outlined previously, such a fixed sample is obtained from the model $P_{\tilde{\theta}}(x)$ with a Metropolis sampler. The method then applies Fisher scoring to find $\theta$ such that $\bar{z}_\theta − z(x_{obs}) = 0$. Since the sample of graphs is an approximation of $\textbf{X}$, a weighted average of the networks statistics is used to determine $\bar{z_{\theta}}$. The sample average $\bar{f}_{\theta} = w^{(1)}f(x^{(1)}) + w^{(2)}f(x^{(2)}) + ... + w^{(M)}f(x^{(M)})$, of a function $f(x)$, with weights \begin{equation} \label{eqn:eqn6} w^{m} = \frac{e^{(\theta_{1} - \tilde{\theta_{1}})z_{1}(x^{(m)}) + (\theta_{2} - \tilde{\theta_{2}})z_{2}(x^{(m)}) + ... + (\theta_{p} - \tilde{\theta_{p}})z_{p}(x^{(m)})}}{\sum_{k=1}^{M}e^{(\theta_{1} - \tilde{\theta_{1}})z_{1}(x^{(m)}) + (\theta_{2} - \tilde{\theta_{2}})z_{2}(x^{(m)}) + ... + (\theta_{p} - \tilde{\theta_{p}})z_{p}(x^{(m)})}} \end{equation} is an asymptotic approximation to the true expected value $E_{\theta}(f(X))$ when $\theta \approx \tilde{\theta}$. To solve the moment equation, a sequence of parameters $\theta^{(0)} = \tilde{\theta}, \theta^{(1)}, \theta^{(2)}, ..., \theta^{(g)}$ is generated using Fisher scoring. The updating step of $\theta^{(g)}$ in the Fisher scoring algorithm is of the form: \begin{equation} \label{eqn:eqn7} \theta^{(g)} = \theta^{(g-1)} - \frac{1}{D(\theta^{(g-1)})} \left\{\sum_{m=1}^{M}w^{m}z(x^{m}) - z(x_{obs}) \right\} \end{equation} and iterates until that $\theta^{(g-1)}$ for which $\sum_{m=1}^{M}w^{m}z(x^{m}) = E_{\theta}(z(X))$ such that $E_{\theta}(z(X)) - z(x_{obs}) = 0$, which is the MLE. The term $D(\theta)$ in the Fisher scoring algorithm is a scaling matrix for the difference between observed values and the expected values of the network statistics. In short, the network statistics need to be scaled because they differ in their sensitivity to changes in the parameter vector $\theta$. The matrix $D(\theta)$ is set to the weighted sample covariance $\sum_{m}w^{(m)}z(x^{(m)})z(x^{(m)})^{T} - $ \\ \noindent $\left[ \sum_{m}w^{(m)}z(x^{(m)}) \right] \left[ \sum_{m}w^{(m)}z(x^{(m)}) \right]^{T}$. 

@@

Having introduced the exponential random graph model (ERGM) for a single observed social network, let us now define the ERGM at the level of the network sample. One method for obtaining an ERGM over a network sample is a meta-analysis \cite{lubbers2003group}. The idea behind a meta-analysis is to use a weighted average to summarize over the results of multiple statistical models, here ERGMs \cite{lubbers2003group}. To illustrate the method, let us imagine a scenario where we want to apply an ERGM to a network sample of size $\{2, ..., m\}$. \clearpage \noindent Initially, an ERGM is fitted to each network in the sample, resulting in a set of ERGMs with size equal to $m$, each with its own set of parameters and standard errors. In order to summarize these separate ERGMs, a meta-analytic approach as outlined in \citeA{lubbers2003group} can be applied. This method splits the coefficients of the separate ERGMs into an average coefficient and a network-dependent deviation \cite{lubbers2003group}. Let $\theta$ denote a parameter included in each ERGM. The meta-level regression equation for this parameter can then be written as: \begin{equation} \label{eqn:eqn8} \hat{\theta} = \mu_{\theta} + U_{m} + E_{m} \end{equation} where $\hat{\theta}$ denotes the estimated parameter value for network $m$, $\mu_{\theta}$ is the average parameter coefficient in the population, $U_{m}$ is the true deviation of network $m$, which has a mean of 0 and a variance $\sigma^{2}_{\theta}$, and $E_{m}$ denotes the estimation error associated with the true parameter value $\theta_{m}$. The variance of $E_{m}$ is assumed to be known and can be obtained as the squared standard error of $\theta_{m}$ \cite{lubbers2003group}. The meta-analysis consequently seeks to differentiate between the true and error variance, in order to obtain more precise estimators for $\mu_{\theta}$ and $\sigma^{2}_{\theta}$. Put differently, by differentiating between true and error variances the method more accurately weights $\mu_{\theta}$ based on differences in standard errors between networks. \citeA{lubbers2003group} suggests using iteratively re-weighted least squares (IRLS) for estimating \eqref{eqn:eqn8}, such that networks with large standard errors have less influence on $\mu_{\theta}$ than networks with small standard errors. This results in a weighted vector of parameter means, each with an associated standard error. These results can then be interpreted as representing a general ERGM description of the network sample. The reader is referred to \citeA{lubbers2003group} for a complete overview and practical example of the meta-analytic procedure.

@@

\clearpage As stated earlier, the information criteria formulated by \citeA{akaike1974new} - Aikake's information criterion (AIC) - and \cite{schwarz1978estimating} - Schwarz's information criterion (SIC) - will here be used to quantify GOF. The AIC and SIC can be defined respectively as \begin{equation} \label{eqn:eqn9} \text{AIC(M)} = -2 \text{log} P_{\hat{\theta}} + 2p, \end{equation} \begin{equation} \label{eqn:eqn10} \text{SIC(M)} = -2\text{log}P_{\hat{\theta}} + \text{log}S. \end{equation} Here, M refers to a particular ERGM, $P_{\bar{\theta}}$ refers to the maximized likelihood under M, $p$ refers to the number of parameters in the ERGM, and $S$ refers to the size of the sample \cite{wit2012all}. For a derivation of these two formulae with respect to the Kullback–Leibler divergence the reader is referred to \citeA{wit2012all}. 
